---
title: "7: Part 2 - Generalized Linear Models"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Apply special cases of the GLM (ANOVA) to real datasets
2. Interpret and report the results of ANOVA in publication-style formats

## Set up
```{r, message = FALSE}

library(tidyverse)
#install.packages("agricolae")
library(agricolae)   #you may need to install/update package "htmltools" before installing and loading agricolae
library(here)
here()

Litter <- read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)

# Set date to date format
Litter$collectDate <- as.Date(Litter$collectDate , format = "%Y-%m-%d")

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

### One-way ANOVA
A one-way ANOVA is used when you have one categorical explanatory variable with more than two categories. In R, we can run the model with the function `lm` or `aov`, the latter of which will allow us to run post-hoc tests to determine pairwise differences.

**For example: Does total litter biomass differ among sites at the Niwot Ridge LTER?**
```{r}
# Wrangle the data
Litter.Totals <- Litter %>%
  group_by(plotID, collectDate, nlcdClass) %>%
  summarise(dryMass = sum(dryMass))

summary(Litter.Totals)
#check all value of plotID
summary(Litter.Totals$plotID)

# Test for normality. -- check our assumptions (normal distribution is 1st)
# Note: the important assumption for GLMs is normality of residuals, 
# not necessarily normality of raw data. See below when we plot the model.
# shapiro.test() performs the Shapiro-Wilk test of the null that data follows a normal distribution
#p-value is less than 0.05 - reject the null (meaning data is not normally distributed), except for 2 (057, 046) - because if p is greater than 0.05 means that they have normal distribution 

shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_040"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_041"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_046"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_047"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_051"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_057"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_058"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_061"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_062"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_063"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_064"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_067"])

#results: reject null in all except two: NIWO_057 and NIWO_046
#but method is robust from small deviations from normal distribution: Q-Q plot shows that data is not normally distributed  
qqnorm(Litter.Totals$dryMass); qqline(Litter.Totals$dryMass)

# Test for equal variance (2nd assumption)
# ANOVA is robust against departures from equal variance.
# bartlett.test() performs Bartlett's test of the null that the variances in each of the groups (samples) are the same.
bartlett.test(Litter.Totals$dryMass ~ Litter.Totals$plotID)

#results: reject null i.e. variances are not equal

# Format ANOVA as aov
#assign obj and call function aov() - similar to lm(), specify dataset and which is continuous variable and which is categorical variable 
Litter.Totals.anova <- aov(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova)

#results: reject null hypothesis i.e. difference between a pair of group means is statistically significant
  #null -- means are the same across the populations; rejecting this but we don't know which means are different from each other (or which pair of differences is significant across means --> post-hoc test can help with this!)

# Format ANOVA as lm - main difference between these two functions is outputs (lm will have row for each plotID) 
Litter.Totals.anova2 <- lm(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova2)

# Checking model fit and assumptions
# ANOVA is robust against departures from normality.
plot(Litter.Totals.anova2)

# Post-hoc test
# TukeyHSD() computes Tukey Honest Significant Differences
TukeyHSD(Litter.Totals.anova)

#diff (difference between means of two sites)
#lwr, upr (lower, upper limits)
#pairwise comparison 

# Extract groupings for pairwise relationships
#call HSD.test(anova object, our variable, group = TRUE -- group by groups with same mean)
#classifying data for you 
Litter.Totals.groups <- HSD.test(Litter.Totals.anova, "plotID", group = TRUE)
Litter.Totals.groups

# Graph the results - generate a plot where x-axis is site info, y-axis is drymass, as boxplot for each siteID, add treatment groups 
#can see differences in means 
Litter.Totals.plot <- ggplot(Litter.Totals, aes(x = plotID, y = dryMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("abc", "ab", "abcd", "bcd", "bcd", "a", 
                         "d", "abcd", "cd", "d", "abcd", "abcd")) +
  labs(x = "Plot ID", y = "Dry Mass (g)") +
  ylim(0, 85)
print(Litter.Totals.plot)
```

What information does the output give us? How might we report this information in a report?

> We would need to reorder the sites, have the same order as the list is showing from the HSD test. This would make the data better represented. 

## TWO-WAY ANOVA (examine two categorical variables)

### Main effects
A two-way ANOVA allows us to examine the effects of **two categorical explanatory variables on a continuous response variable.** 

For example: What if we wanted to know how litter masses varied across functional groups and land cover class?
```{r}
# Format as aov
Litter.anova.2way <- aov(data = Litter, dryMass ~ functionalGroup + nlcdClass)
summary(Litter.anova.2way)
summary(Litter$nlcdClass)
summary(Litter$functionalGroup)

# Format as lm - intercept is with respect to one of levels
Litter.anova.2way2 <- lm(data = Litter, dryMass ~ functionalGroup + nlcdClass)
summary(Litter.anova.2way2)

# Run a post-hoc test for pairwise differences - all possible combos between groups, means, and confidence intervals (you will be able to extract more info than ANOVA)
TukeyHSD(Litter.anova.2way)

```

### Interaction effects

We may expect the effects of functional group and land cover to be dependent on each other. For instance, we may expect needle biomass to be larger in evergreen forests than in other land types. In this case, we might choose to run an **interaction effects two-way ANOVA**, which will examine the individual effects of the explanatory variables as well as the interaction of the explanatory variables.

The output gives test statistics for each explanatory variable as well as the interaction effect of the explanatory variables. **If the p-value for the interaction effect is less than 0.05, then we would consider the interaction among the explanatory variables to be significant.**

```{r}
# Format as aov
#instead of dryMass as a function of group and class, it is based on interactions?
Litter.anova.2way3 <- aov(data = Litter, dryMass ~ functionalGroup * nlcdClass)
summary(Litter.anova.2way3)

# Format as lm
Litter.anova.2way4 <- lm(data = Litter, dryMass ~ functionalGroup * nlcdClass)
summary(Litter.anova.2way4)

```

If the interaction is significant, we interpret pairwise differences for the interaction. If the interaction is not significant, we interpret differences for the main effects only.

```{r}
# Run a post-hoc test for pairwise differences
TukeyHSD(Litter.anova.2way3)

Litter.interaction <- with(Litter, interaction(functionalGroup, nlcdClass))

#Litter.test <- cbind(Litter,Litter.interaction)

#compair means of each litter interactions for 25 levels found from Tukey
Litter.anova.2way5 <- aov(data = Litter, dryMass ~ Litter.interaction)

Litter.groups <- HSD.test(Litter.anova.2way5, "Litter.interaction", group = TRUE)
Litter.groups
#grouped dataset based on interactions 

# Plot the results
# Edit this graph to make it more visually appealing.
# Add letters to represent statistical groupings.
Litter.anova.plot <- ggplot(Litter, aes(y = dryMass, x = nlcdClass, color = functionalGroup)) +
  geom_boxplot()
print(Litter.anova.plot)

#make graph look pretty with labels 
```

