---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Meg Manning"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1 set up session
library(tidyverse)
library(agricolae)
library(lubridate)
library(here)
here()

#read in raw data
NTL_LTER <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
  stringsAsFactors = TRUE)
# Set date to date format
NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledate, format = "%m/%d/%y")
class(NTL_LTER$sampledate)

#2 build a ggplot theme
mytheme <- theme_light(base_size = 14) +
  theme(axis.text = element_text(color = "maroon"),
        legend.position = "right",
        plot.background = element_rect(fill = "lightblue"))
theme_set(mytheme)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean temperature does not change with depth across all lakes
Ha: Mean temperature does changes with depth across all lakes


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4 filter data with pipe function
NTL_LTER.July <- NTL_LTER %>%
  filter(month(sampledate) == 7) %>%
  select(lakename:temperature_C) %>%
  drop_na()

#5 Build a scatter plot of temperature by depth
tempbydepth.plot <- ggplot(NTL_LTER.July,
                            aes(x = depth,
                                y = temperature_C)) +
  geom_point() +
  geom_smooth(method = "lm",
              col = "blue") +
  ylim(0, 35) +
  labs(x = "Depth (m)",
       y = "Temperature (°C)",
       title = "Temperature by Depth")
print(tempbydepth.plot)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: In this plot, we can see that that at lower depths or near the surface, temperature is higher. There seems to be a suggestion of a relationship that with increased depth, there will be a drop in temperature, which seems intuitive. However, it is hard to tell whether the linear model on the graph is the best model for this relationship since there are data points that fall outside of the geom_smooth line in a way that seems like it could be better explained by a different model method.

7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7 Linear regression to test relationship between temp and depth
tempbydepth.regression <- lm(data = NTL_LTER.July, 
                             temperature_C ~ depth)
summary(tempbydepth.regression)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: Our summary statistics show a mean surface temperature of about 22 degrees celcius because the x-intercept (when depth is 0) then the mean temperature is ~22°C. The regression coefficient also shows that if you increase the depth by 1 meter, temperature will decrease by ~2°C. Both p-values are less than 0.05, meaning that this regression is statistically significant and the R-squared value indicates that depth explains ~74% of the temperature. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9 Run and AIC for best variables 
Tempbyall.regression <- lm(data= NTL_LTER.July, 
                            temperature_C ~ depth + year4 + daynum)
step(Tempbyall.regression)

#the AIC value does not decrease when taking away depth, year or daynum
#best regression is with all 3 variables

#10 Run a multiple regression on depth, year4, and daynum variables
tempbest.regression <- lm(data= NTL_LTER.July, 
                           temperature_C ~ depth + year4 + daynum)
summary(tempbest.regression)
step(tempbest.regression)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: Can't take away any of them...



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12 mean temps of different lakes for July - ANOVA
NTL.anova <- aov(data = NTL_LTER.July, temperature_C ~ lakename)
summary(NTL.anova)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Based on the fact that the p-value is less than 0.05, we can reject the null hypothesis that the mean teamperature is the same across all lakes, and we can state that the mean temperature if different for each lake.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14. graph of temp (y) by depth (x) colorized by lakename
tempbydepth.plot2 <- ggplot(NTL_LTER.July,
                            aes(x = depth,
                                y = temperature_C,
                                color = lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm",
              se = FALSE) +
  labs(title = "Temperature by Depth",
       color = "Lake Name",
       x = "Depth (meters)",
       y = "Temperature °C")
print(tempbydepth.plot2)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15 Run Tukey HSD on 
TukeyHSD(NTL.anova)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#Wrangle data for Crampton Lake and Ward Lake
NTL.CramptonWard <- filter(NTL_LTER.July,
                           lakename == "Crampton Lake" | lakename == "Ward Lake")

#Two-sample t-test
NTL.twosample <- t.test(NTL.CramptonWard$lakename ~ NTL.CramptonWard$temperature_C)
NTL.twosample

```

>Answer: 
