---
title: "Assignment 2: Coding Basics"
author: "Meg Manning"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. #create a sequence from 1-30 increasing by 3 and assign name
seq1 <- seq(1,30,3)
View(seq1)

#2. #Compute mean and median of seq1 using mean() and median() functions
mean(seq1)
median(seq1)

#3.#ask if mean is greater than median
mean(seq1) > median(seq1)
#since they are the same, it gives the result FALSE, meaning the mean is not greater than the median
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#create a vector for names using c()
names <- c("Meg", "Josh", "Nick", "Melissa") #character or string vector
#Create a vector for test scores out of 100
test_results <- c(99, 75, 50, 92) #numeric
#create a vector for if they passed
passed_or_not <- c(TRUE, TRUE, FALSE, TRUE) #logical vector

#combine vectors into a data frame and assign a name
student_results <- cbind(names,test_results,passed_or_not)

#create vector of labels
student_results_labels <- c("Student Names", "Test Result of out 100", "If Student Passed")
View(student_results_labels)

#Option 2
student_results <- cbind("Student Names" = names,"Test Result out of 100" = test_results,"If Student Passed" = passed_or_not)

#create labels 
colnames(student_results) <- student_results_labels

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: This data frame is different from a matrix because it contains mutliple classes of data within it, like a spreadsheet. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#create function with if/else statement - if test score above 50 is passing'
#Vector of scores 
test_results

#create function for passing/failing results using ifelse
passing_grade <- function(x){
  print(ifelse(x>50, TRUE, FALSE))
}
passing_grade(test_results)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: ifelse worked better because it was a simple function to use because regardless of what you set the function to do with your input, it will either result in a "TRUE" or "FALSE" statement which is what we are trying to figure out. 
