---
title: "Lab 3 - Data Types | Exploring Datasets"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Fall 2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives

1.  Discuss and navigate different data types in R
2.  Create, manipulate, and explore datasets
3.  Date objects

## Data Types in R

R treats objects differently based on their characteristics. For more information, please see: <https://www.statmethods.net/input/datatypes.html>.

-   **Vectors** 1 dimensional structure that contains elements of the same type.

-   **Matrices** 2 dimensional structure that contains elements of the same type.

-   **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

-   **Lists** Ordered collection of elements that can have different modes. (many functions we used in class with give output as a list) 

-   **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (\# at beginning of line, ---- at end)

```{r Create various data structures}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector
vector1
vector2 <- c("one","two","three") # character vector
vector2
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
vector3

vector1[3] #use square bracket to point to index of element you are looking for (ex. this will be 5.3) 

# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) #must specify number of rows, numbers of columns 
matrix1
matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) #by row = do you want to go by rows or by columns
matrix2
matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE, # return after comma continues the line
                  dimnames = list(c("uno", "dos", "tres", "quatro", "cinco"), 
                                  c("un", "deux", "trois", "quatre"))) #

matrix1[4, ] #4th row of matrix, and all elements in that row*
matrix1[ , 3] #3rd column of matrix and all elements in that column*
matrix1[c(12, 14)] #want the elements in index 12 and index 14
matrix1[c(12:14)] #calling out 12th-14th elements
matrix1[2:4, 1:3] #2nd row, 4th element, 1st column, 3rd element

cells <- c(1, 26, 24, 68)
rnames <- c("R1", "R2")
cnames <- c("C1", "C2") 
matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) # 
matrix4

# Lists ---- 
list1 <- list(name = "Maria", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1
list1[[2]]

# Data Frames ----
d <- c(1, 2, 3, 4) # What type of vector? --> numeric vector
e <- c("red", "white", "red", NA) # What type of vector? --> character vector
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector? --> logic vector
#want to check the type of vector? look at environment and see the abr. before the data (num, chr, logi)

dataframe1 <- data.frame(d,e,f) #create a data frame and put all three vectors together
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) #if you want to rename the columns for three variables you can use this

dataframe1[1:2,] # 
dataframe1[c("ID","Passed")] # 
dataframe1$ID

```

Question: How do the different types of data appear in the Environment tab?

> Answer: The types will show up with abbreviations (num, logi, chr). Matrices and Dataframes show up as data, vectors will show up under values, functions will show up as a third category. 

Question: In the R chunk below, write "dataframe1\$". Press `tab` after you type the dollar sign. What happens?

> Answer: R will suggest all the columns you have in your dataframe, their names, and which one you are trying to access. 

### Coding challenge

Find a ten-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create two vectors, one representing the high temperature on each of the ten days and one representing the low.

```{r DataExercise-1a}
#create 2 vectors with highest and lowest temp in 10-day 
hightemp <- c(81, 83, 83, 84, 78, 71, 79, 82, 85, 84)
lowtemp <- c(55, 56, 58, 60, 61, 63, 63, 64, 64, 63)
```

Now, create two additional vectors that include the ten-day forecast for the high and low temperatures in Celsius. Use a function to create the two new vectors from your existing ones in Fahrenheit.

```{r DataExercise-1b}
#create 2 vectors where temp is converted from Fahrenheit to Celsius
celsius <- function(x) {return((x-32)/(9/5))
  }
hightemp_c <- celsius(hightemp)
hightemp_c

lowtemp_c <- celsius(lowtemp)
lowtemp_c
```

Combine your four vectors into a data frame and add informative column names.

```{r DataExercise-1c}
#combine 4 vectors into dataframe with informative column names 
df_temp <- data.frame("High Temp in F" = hightemp, "High Temp in C" = hightemp_c, "Low Temp in F" = lowtemp, "Low Temp in C" = lowtemp_c)

```

Use the common functions `summary` and `sd` to obtain basic data summaries of the ten-day forecast. How would you call these functions differently for the entire data frame vs. a single column? Attempt to demonstrate both options below.

```{r DataExercise-1d}
#Use summary function to determine summary statistics
summary(df_temp)
#use function sd (standard deviation) for 10-day forecast 
#remember the input can't be a dataframe it must be a list
sd(df_temp$High.Temp.in.F)
```

### Date objects

Remember formatting of dates in R: (can use as.date(), but maybe more useful is *ymd()* or *ydm()* or *mdy*() - doesn't matter if there are abr. years or months - to change from factor as an object)

%d day as number (0-31) %m month (00-12, can be e.g., 01 or 1) %y 2-digit year %Y 4-digit year %a abbreviated weekday %A unabbreviated weekday %b abbreviated month %B unabbreviated month

```{r Coding with dates}
# Adjust date formatting for today
# Write code for three different date formats. 
# An example is provided to get you started.
# (code must be uncommented)
today <- Sys.Date()
format(today, format = "%B")
#format(today, format = "%b")
#format(today, format = "%Y")
#format(today, format = "%y")

```

### Package lubridate

Install and load the package lubridate into your R session. Lubridate offers fast and user friendly parsing of date-time data. Create a string for today's data and then convert it to R date object using lubridate.

More info on lubridate [here][<https://cran.r-project.org/web/packages/lubridate/lubridate.pdf>].

```{r DateExercise1}
#install.packages("lubridate")
library(lubridate)

#Ex1
str_today <- "2023-feb-7"
#Since the format is year-month-day we will use function ymd()
date_obj_today <- ymd(str_today) #will show up in env. without "" meaning it's numerical now
date_obj_today

#Ex2
str_today2 <- "Feb 7, 2023"
#Since the format is month-day-year we will use function mdy()
date_obj_today <- mdy(str_today2)
date_obj_today

#Ex_3 - on your own...
str_juneteenth <- "19 June 1865"
#Since the format is day-month-year we will use function dmy()
date_juneteenth <- dmy(str_juneteenth)
date_juneteenth
```

```{r DateExercise2}
#century issue
#if year is only 2 digits, it might not be the correct century
str_past <- "55-feb-3"
date_obj_past <- ymd(str_past)
date_obj_past

#Build a function to fix year that is more general than the one discussed in the lesson
#function will show up below values
fix.early.dates <- function(d, cutoff) {
       m <- year(d) %% 100  #operator %% is a modular division i.e. integer-divide year(d) by 100 and returns the remainder 
       year(d) <- ifelse(m > cutoff, 1900+m, 2000+m)  #this will update year(d), year() is a function that returns the year for a data object
       return(d)
}

fixed_date_obj_past <- fix.early.dates(date_obj_past,cutoff=23) #cutoff could be the current year to be more general or any other depending on data set 
fixed_date_obj_past
```

```{r centuryfix}
#Fix for century issue
str_past <- "55-feb-3"
#Alternative 1 - fast_strptime(), must specify format, and cutoff date
date_obj_past <- fast_strptime(str_past,"%y-%b-%d",cutoff_2000=23L)
date_obj_past

#Alternative 2 - use parse_date_time2() cutoff is the current year we are in
date_obj_past2 <- parse_date_time2(str_past,"ymd",cutoff_2000=23L)
date_obj_past2

#Functions ymd(), mdy(), ydm() do not take argument cutoff_2000
```

In some cases when dates are provided as integers, you may need to provide an origin for your dates. For example, excel date could be given as number of days since an origin date. Origin date can be different. When R looks at dates as integers, its origin is January 1, 1970. Check if that is true on your machine.

```{r dateorigin}
#Check if "1970-01-01" is your origin date.
lubridate::origin

```
